Gut-resident microorganisms and their genes are associated with cognition and neuroanatomy in children

### 2023-09-01 manuscript/main.tex
write up fig 2 subscales

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive ability score (Table 2, Figure 2C-G) and found that
several metabolic pathways were either significantly enriched or depleted in
children with higher cognitive ability scores. This was true both when
considering all ages (0-120 months) together, though the enrichment of most
pathways was more pronounced either in children under 6 months or those over 18
months. For example, genes for degrading the 3-carbon SCFA, propionate,
were significantly depleted in children with higher cognitive ability
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q = 0.052).
In children over 18 months old, synthesis of menaquinone (vitamin K)
was also negatively associated with cognitive function score (menaquinone synthesis I,
E.S. = -0.170, q = 0.0183), while
genes for the synthesis of the amino acids
glutamate and tryptophan were significantly enriched in children with
higher cognitive function scores (glutamate synthesis I, E.S. = 0.242, 
q = 0.047; tryptophan synthesis, E.S. = 0.119, q = 0.041).
Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.

>>>>

Each of the cognitive assessments (MSEL, WPPSI, WISC)
are composites of multiple subscales that assess different cognitive domains.
While the composite scores are broadly analagous accross ages,
the subscales assessed in each test are different.
In order to assess whether microbial taxa were associated

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive ability score (Table 2, Figure 2D-F) and found that
several metabolic pathways were either significantly enriched or depleted in
children with higher cognitive ability scores. This was true both when
considering all ages (0-120 months) together, though the enrichment of most
pathways was more pronounced either in children under 6 months or those over 18
months. For example, genes for degrading the 3-carbon SCFA, propionate,
were significantly depleted in children with higher cognitive ability
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q = 0.052).
In children over 18 months old, synthesis of menaquinone (vitamin K)
was also negatively associated with cognitive function score (menaquinone synthesis I,
E.S. = -0.170, q = 0.0183), while
genes for the synthesis of the amino acids
glutamate and tryptophan were significantly enriched in children with
higher cognitive function scores (glutamate synthesis I, E.S. = 0.242, 
q = 0.047; tryptophan synthesis, E.S. = 0.119, q = 0.041).
Taken together, these results suggest that
microbial metabolic activity, particularly the metabolism (synthesis and
degradation) of neuroactive compounds may have effects on cognitive
development.
